# General Information {.unnumbered}

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = FALSE)
library(webshot2)
library(downlit)
library(xml2)
library(kableExtra)
source("R/helpers.R")
options(knitr.kable.NA = '')
``` 


```{r results='asis'}
"Course name:	From Patterns to Principles: Theory Construction in the Social Sciences
Instruction language:	English
Type of instruction:	Interactive lectures and group work 
Level:	Excellence Program TSB
Course load:	6 ECTS
Lectures:	Dr. C. van Lissa"->tmp

tab <- structure(c("Course name", "Instruction language", "Type of instruction", 
"Level", "Course load", "Lectures", "From Patterns to Principles: Theory Construction in the Social Sciences", 
"English", "Interactive lectures and group work", "Excellence Program TSB", 
"6 ECTS", "Dr. C. J. Van Lissa"), dim = c(6L, 2L))
knitr::kable(tab, col.names = NULL)
```

This course combines literature from empirical social science, methods and statistics, and philosophy to offer a multidisciplinary perspective on theory in the social sciences. There is no textbook: students will read and present on primary literature and discuss it in class. The course uses challenge based learning and a flipped classroom. While default challenges and corresponding materials are available, students are encouraged to bring their own, thereby connecting this course to their other coursework (or future thesis). The course relies on group work and explicitly allows for specialization (i.e., not all students need to master all course-relevant skills).


## Learning objectives

This course introduces students to the foundations and applications of theory in social science research. We will analyze the need for better theories based on the current "theory crisis" in empirical social science. We define the characteristics and affordances of "good theories": what do they look like, and how do we use them in (empirical) research? Students will engage with theory construction methods to formalize existing theories or construct their own, and are introduced to formal modeling and simulation studies as tools for theory development and testing. By the end of the course, students will be able to critically assess existing theories and their usage in empirical articles, construct theory, simulate theory-implied data, use theory in empirical research, and relate critically to the state of theory in the social sciences.

## Skills

After taking this course, students are able to:

* Remember relevant constructs from the literature
* Understand theory construction methods and their philosophical pitfalls
* Apply theory in empirical research
* Create new - or formalize existing - theory
* Evaluate the use of theory in empirical research

## Study Load

This is the breakdown of the expected study load:

```{r, results='asis'}
tab_hours <- readODS::read_ods("hours_breakdown.ods")
make_table(tab_hours)
```

## Assessment

Your grade for the course is based entirely on your "portfolio", comprised of the following components:

```{r, results='asis'}
tab_ass <- read.csv("assessment.csv")
make_table(tab_ass)
```

## Group Work

For this course, we want to stimulate you to engage with the material, with one another, and with the teachers. Therefore, some portfolio assignments are made in groups.
There are aspects of learning in groups that can really improve your knowledge, like discussion and peer feedback.
Groups comprise 3-4 members and are created during the first tutorial session (@sec-tutorialgroup).
We encourage students to reflect on their [roles in group work](https://uwaterloo.ca/centre-for-teaching-excellence/catalogs/tip-sheets/group-work-assignment-roles), and to explicitly divide roles before each session or switch roles each week.
At minimum, it's useful to always have one Facilitator. If your group tends to be unanimous, appoint a Devil's advocate. If your group tends towards disagreement, appoint an Arbitrator.

### Course materials

**You do not need a book for this course.** This GitBook provides all relevant course materials,
along with primary literature.

### Course schedule 

Always check the latest course schedule on [TimeEdit](https://www.timeedit.net/).
For an overview of the content, see below:

```{r, results = "asis", echo = FALSE}
library(DT)
schedule <- readODS::read_ods("schedule2.ods")
schedule <- schedule[rowSums(is.na(schedule)) < 12, c("Dates", "Activities", "Topic", "Assessment", "Homework")]

make_table(schedule)

```


### Communication

Communication about the course occurs through Canvas (Login with your student ID and password).



## Staff

**Coordinator/lecturer:**

[dr. Caspar J. Van Lissa](https://www.tilburguniversity.edu/staff/c-j-vanlissa)

**Lab sessions**

TBD
<!-- [Danielle McCool](https://www.uu.nl/staff/DMMcCool) -->
<!-- [Laura Hofstee](https://www.uu.nl/staff/LHofstee) -->



## Contact hours

This course consists of 14 lectures and 11 lab sessions (LA).
<!-- Attendance is not mandatory (this applies to both the lectures and lab sessions), but experience shows that it is very unlikely that students who do not attend will pass the course. -->

The course consists of a common part and a major specific part. The first 10 lectures are common lectures, which means they have to be followed by students from all majors (i.e., business and economics, cognitive neuroscience and social sciences). The last four lectures are major specific. 

During the lab sessions you will work on applied exercises supervised by a teaching-assistant. These exercises are always related to topics discussed during the lectures.
The assignments of the SPSS lab sessions are available via Canvas. You will be able to ask your lab teacher questions during the lab sessions. 


## Software

The official software for this course is SPSS, which is available on the university computers. However, there are many free alternatives to SPSS. If you want to use your own computer, you might consider trying one of these:

*	PSPP, which is designed to be nearly identical to SPSS with all the same basic functionality: https://www.gnu.org/software/pspp/pspp.html
* JASP, which is more modern, looks nicer and is very easy to use – but less similar to SPSS: 
https://jasp-stats.org/

