# Setup your Computer {.unnumbered}


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This is a tutorial on how to set up your personal computer for the course.
It guides you through the installation of several software packages.
This vignette does not assume a prior installation of `R`, so it is suitable for novice users.
You only have to perform these steps once for every computer you intend to use,
and the entire process should take approximately 30 minutes if you start from scratch.
In case some of the software is already installed on your system, you can skip those related steps.

Follow these steps in order:

1. Install [R (free)](https://CRAN.R-project.org)
1. Install ['RStudio' Desktop (Free)](https://posit.co/download/rstudio-desktop/)
1. Open RStudio. Once it's loaded, you'll see a panel in the bottom left labeled "Console". Click there, then copy-paste the code block below this writing, line by line, and press enter. Be prepared for three contingencies:  
    + If you receive any error saying *There is no package called [package name]*, then run the code `install.packages("package name")`  
    + If you are prompted to update packages, just press [ENTER] to avoid updating packages. Updating packages this way in an interactive session sometimes leads to errors if the packages are loaded.  
    + If you see a pop-up dialog asking *Do you want to install from sources the package which needs compilation?*, click *No*.


```{r eval = FALSE, echo = TRUE}
install.packages("theorytools", dependencies = TRUE)
install.packages("dagitty")
```
